Missing The Point

### 2020-11-13 Manuscript.Rmd
Comments Gerko 12/11

Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences. A popular technique to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset. Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18]. To obtain imputations, most imputation software packages rely on iterative algorithms.

With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained using the imputed data? These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied (Van Buuren, 2018, 6.5.2). While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@liu14; @zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence. 
<!-- (Zhu & Raghunathan, 2015; Takahashi, 2017) -->

@@

In this paper, we explore diagnostic methods for iterative imputation algorithms. For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. We consider two non-convergence identifiers for iterative algorithms: autocorrelation (conform Lynch, 2007, p. 147) and potential scale reduction factor $\widehat{R}$ (conform Vehtari et al., 2019, p. 5). Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence in the multivariate state-space of the algorithm.

We propose a novel multivariate parameter to check for non-convergence in iterative algorithms. We aim to show which method is the most informative about non-convergence in iterative imputation [explain that method = parameter + diagnostic + interpretation].

@@

    for each simulation repetition (1:1000)
      1. simulate complete data
      for each missingness condition (1:9)
        2. create missingness
        for each iteration (1:50)
          3. impute missingness
          4. perform analysis of scientific interest
          5. apply non-convergence identifiers
          6. pool results across imputations
          7. compute performance measures
        8. combine outcomes from all iterations
      9. combine outcomes from all missingness conditions
    10. aggregate outcomes from all simulation repetitions

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].

@@

[Add mvtnorm package here??] The complete set is amputed according to nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases. [Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences). And add the `ampute` function?]

@@

$$
Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
$$
where $Y'$ is the expected value of the outcome. Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $r^2$ that we obtain after pooling the regression results across the imputations.

@@

As recommended by @buur18, our performance measures are bias ($E(\bar{Q})-Q$), average confidence interval width ($E(\bar{Q}_{LL} - \bar{Q}_{UL})$, where $LL$ and $UL$ denote the lower and upper limit of the 95% confidence interval respectively) and empirical coverage rate ($Pr(\bar{Q}_{LL} \geq Q \geq \bar{Q}_{UL})$) of the estimands [or just of the regression estimate?? otherwise add the ciw and cov of r2 too!! and check if definition of ciw is correct].

@@

The following figures display the simulation results for each method and each performance measure, contrasted to the number of iterations in the imputation algorithm. Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].

@@

We observe that autocorrelation in the chain means rapidly decreases until $it \geq 6$. This means that there is some initial trending within chains, but stationarity of the average imputed value does not improve substantively after the first few iterations. These results hold irrespective of the missingness condition.

@@

With this study, we show that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached. Any further iterations would then burn computational resources without improving the statistical inferences.

@@

- The figures show results of a `mice` run with 30 iterations but otherwise default settings.

@@

From the traceplot of the chain means it seems that mixing improves up-to 20 iterations, and there is some trending up-to 20 iterations as well. The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations.

@@

[Is this a nice example? Why is convergence so slow? Don't the defaults work? Is this the bmi+wgt+wgt problem? Maybe find an easier problem?]

>>>>

Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences. The de facto standard to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset. To generate imputed values, most imputation software packages rely on iterative algorithms. [Why iterative? We estimate the posterior predictive distribution of the missing values; our estimates become better as we go; there should be a point at which our estimates are non-improving.] Iterative imputation techniques have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18].

With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained using the imputed data? These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. [Common convergence diagnostics may not work.] While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@liu14; @zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence. 
<!-- (Van Buuren, 2018, 6.5.2) (Zhu & Raghunathan, 2015; Takahashi, 2017) -->

@@

In this paper, we explore some existing and new means of diagnosing non-convergence in iterative algorithms. We evaluate the validity of these methods in the context of missing data imputation using model-based simulation in `R` [@R]. [Brief overview of paper: the diagnostics, the simulation, the results, the discussion, a case study.]

<!-- In this paper, we explore diagnostic methods for iterative imputation algorithms. For reasons of brevity, we focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. We consider two non-convergence identifiers for iterative algorithms: autocorrelation (conform Lynch, 2007, p. 147) and potential scale reduction factor $\widehat{R}$ (conform Vehtari et al., 2019, p. 5). Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence in the multivariate state-space of the algorithm. -->

<!-- We propose a novel multivariate parameter to check for non-convergence in iterative algorithms. We aim to show which method is the most informative about non-convergence in iterative imputation [explain that method = parameter + diagnostic + interpretation]. -->

@@

    for each simulation repetition 
      1. simulate complete data
      for each missingness condition 
        2. create missingness
        for each iteration 
          3. impute missingness
          4. perform analysis of scientific interest
          5. apply non-convergence identifiers
          6. pool results across imputations
          7. compute performance measures
        8. combine outcomes from all iterations
      9. combine outcomes from all missingness conditions
    10. aggregate outcomes from all simulation repetitions

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].

@@

The complete set is amputed according to nine missingness conditions. We use a $3 \times 3$ factorial design consisting of three missingness mechanisms and three proportions of incomplete cases. [Briefly describe MCAR, MAR, MNAR + 25, 50, 75% of cases incomplete (not using 5% anymore, because according to Vink, n.d., 25% is more representative of missing data problems in the social sciences). And add the `ampute` function?]

@@

$$
\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2,
$$
where $\hat{Y}$ is the predicted value of the outcome. Our estimands are the regression coefficient $\beta_1$ and coefficient of determination $\rho^2$ that we obtain after pooling the regression results across the imputations.

@@

As recommended by @buur18, our performance measures are bias, confidence interval width, and coverage rate of the estimands ($\S$ 2.5.2)[or just of the regression estimate?? otherwise add the ciw and cov of r2 too!!].

@@

The following figures display the simulation results for the eight diagnostic methods and four performance measures, contrasted to the number of iterations in the imputation algorithm. Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors]. Note that these results are averages of the $n_{sim} = 1000$ simulation repetitions.

@@

Autocorrelation in the chain means decreases rapidly in the first few iterations. The decrease is substantive until $it \geq 6$. This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity. These results hold irrespective of the missingness condition.

@@

The validity of iterative imputation stands or falls with algorithmic convergence---or so it's thought. We have shown that iterative imputation algorithms can yield correct outcomes even when a converged state has not yet formally been reached. Any further iterations just burn computational resources without improving the statistical inferences.

@@

- The figures show results of a `mice` run with 20 iterations but otherwise default settings.

@@

From the traceplot of the chain means it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.

<!-- and there is some trending up-to 20 iterations as well. The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations. -->

@@
